---
title: "Identify long pass and 3pt shooting"
output: 
  pdf_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(dplyr)
library(gganimate)
source("normal-hmm-2state.R")
```

# Simply check the movement and event data

```{r, echo = FALSE}
event <- read_csv('data/CHA-TOR-events.csv', col_types = cols())
data <- read_csv("data/CHA-TOR.csv", col_types = cols())
head(data)
head(event)
```

# Look at Event 1 and the movement of ball

There seems to exist two pass in the ball movement. Kyle Lowry initially dribbles the ball and pass it to others

```{r, echo = FALSE}
defaultW <- getOption("warn")
options(warn = -1)
event1 <- read_csv('data/event1_passing.csv', col_types = cols()) %>% rename(index = 'X1')
event1_ball <- event1 %>% filter(player_id == 'ball')
event1 %>% filter(player_id == 'ball') %>% ggplot() + aes(x = x_loc_original, y = y_loc_original, color = player_id) %>% geom_point() + xlim(0, 100) + ylim(0, 50) + ggtitle("Movement of the ball")
event1 %>% filter((player_id == 'ball' | player_id == 'Kyle Lowry' | player_id == "DeMar DeRozan") & game_clock > 708.5) %>% ggplot() + aes(x = x_loc_original, y = y_loc_original, color = player_id) %>% geom_point() + xlim(0, 100) + ylim(0, 50) + ggtitle("The first pass and players")
event1 %>% filter((player_id == 'ball' | player_id == "DeMar DeRozan" | player_id == "Luis Scola") & game_clock < 708.5 & game_clock > 706.5) %>% ggplot() + aes(x = x_loc_original, y = y_loc_original, color = player_id) %>% geom_point() + xlim(0, 100) + ylim(0, 50) + ggtitle("The second pass and players")
```

# Fit the passing

We preprocess the data (using python) and calculate the speed for each player and the ball at each timestep.

```{r, echo = FALSE}
head(event1_ball)
```

```{r, echo = FALSE}
plot(event1_ball$speed, type = 'l')
```

After investigating the speed data for these two potential state, we assign the initial value and starting fitting two states, a pass occurs or a pass doesn't occur.

```{r}
# number of states, 2
m <- 2
# the mean of two state
mu <- c(9, 26)
# standard deviation of two states
sigma <- c(12, 4)
# transition probability
gamma <- matrix(c(0.99, 0.01, 0.02, 0.98), m, m, byrow = TRUE)
# Initial state probability
delta <- c(1, 0)
```

```{r, echo = FALSE}
defaultW <- getOption("warn")
options(warn = -1)
# Fit the hmm model by maximum likelihood
event1_ball_speed <- norm.HMM.fit(event1_ball$speed, m, mu, sigma, gamma, stationary = TRUE)
# Calculate the se for the model params
event1_ball_speed_se <- norm.HMM.params_SE(event1_ball$speed, 20, event1_ball_speed, stationary = TRUE)
event1_ball_speed
```

```{r, include = FALSE}
defaultW <- getOption("warn")
options(warn = -1)
ball_speed_range <- round(min(event1_ball$speed)):round(max(event1_ball$speed))
ball_speed_ci_plot <- norm.HMM.CI_MonteCarlo(ball_speed_range, m = 2, n = length(event1_ball$speed), event1_ball_speed_se)
state_seq <- norm.HMM.viterbi(event1_ball$speed, event1_ball_speed)
event1_ball_speed$GuessState <- as.factor(state_seq)
```

```{r, echo = FALSE}
data_ball_speed <- data.frame(Time = c(1: length(event1_ball$speed)),
                              Observation = event1_ball$speed,
                              x_loc_original = event1_ball$x_loc_original,
                              y_loc_original = event1_ball$y_loc_original)
data_ball_speed$GuessState <- as.factor(state_seq)
data_ball_speed <- data_ball_speed %>% mutate(smooth_speed = ifelse(GuessState == 1, 10, 35))
norm_hist_dist_CI(m, data_ball_speed, event1_ball_speed, ball_speed_ci_plot, width = 1)
timeseriesfit(head(data_ball_speed, 150))
```

```{r, include = FALSE}
# gif_passing <- data_ball_speed %>% ggplot() + aes(x = x_loc_original, y = y_loc_original, color = GuessState, size = smooth_speed) + geom_point() + xlim(0, 100) + ylim(0, 50) + transition_time(Time) + ease_aes('linear')
# animate(gif_passing, renderer=gifski_renderer("event1_passing.gif"))
```


# 3-pt shooting

We also prepare the data for three point shooting.

```{r, echo = FALSE}
defaultW <- getOption("warn")
options(warn = -1)
event28 <- read_csv("data/event28_3pt.csv", col_types = cols())
event28 <- event28 %>% rename(index = 'X1')
three_pt <- event28 %>% filter(game_clock > 549)
head(three_pt)
```

I want to fit it together with passing/dribbling, but right now I think the speed enough for long passing, but it is not enough to identify the dribble by player and shoot by player (since they are close). 

I believe distance to the hoop could be another criteria for us to identify a shooting.  

```{r, echo = FALSE}
three_pt %>% filter(team_id == 'ball' & game_clock > 554.5) %>% ggplot() + aes(x = x_loc_original, y = y_loc_original) %>% geom_point() + xlim(0, 100) + ylim(0, 50)
plot((three_pt %>% filter(team_id == 'ball'))$speed, type = 'l')
```

Lastly, I create two gif plots.

```{r, include = FALSE}
# gif_3pt <- three_pt %>% filter(team_id == 'TOR' | team_id == 'ball') %>% ggplot() + aes(x = x_loc_original, y = y_loc_original, color = player_id, size = speed) + geom_point() + xlim(0, 100) + ylim(0, 50) + transition_time(index) + ease_aes('linear')
# animate(gif_3pt, renderer=gifski_renderer("event28_3pt.gif"), duration = 20)
```








